---
title: "Interactive Fixed Effects for Balanced and Unbalanced Panels"
author: "Binzhi Chen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Interactive Fixed Effects for Balanced and Unbalanced Panels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4
)
```

# Introduction

## Why interactive fixed effects?

Standard two-way fixed effects (TWFE) controls for unobserved heterogeneity
through an additive structure:

$$y_{it} = \alpha_i + \xi_t + X_{it}'\beta + u_{it}$$

The unit effect $\alpha_i$ absorbs any time-invariant unit characteristic;
the time effect $\xi_t$ absorbs any unit-invariant time shock.  This works
well when heterogeneous responses to common shocks are negligible.

Many empirical settings violate this assumption.  Consider:

- **Globalisation shocks**: countries are exposed differently depending on
  their trade openness (a unit characteristic that interacts with a common
  world-demand shock).
- **Monetary policy**: firms with different financial leverage respond
  asymmetrically to a common interest-rate change.
- **COVID-19**: regional outcomes depend on a national shock weighted by
  pre-existing health-system capacity.

In each case, the unobserved confounder has the product form
$\lambda_i' F_t$: a unit-specific loading $\lambda_i$ multiplied by a
common factor $F_t$.  If $X_{it}$ is correlated with $\lambda_i' F_t$,
TWFE is inconsistent even as $N, T \to \infty$.

## The IFE model

Bai (2009) proposes replacing the additive error structure with:

$$y_{it} = \alpha_i + \xi_t + X_{it}'\beta + \lambda_i' F_t + u_{it}
\qquad (1)$$

where:

- $\beta \in \mathbb{R}^p$: structural parameters of interest,
- $F_t \in \mathbb{R}^r$: common factors ($T \times r$ matrix $\hat{F}$,
  normalised $\hat{F}'\hat{F}/T = I_r$),
- $\lambda_i \in \mathbb{R}^r$: unit-specific factor loadings
  ($N \times r$ matrix $\hat{\Lambda}$),
- $u_{it}$: idiosyncratic error with $\mathbb{E}[u_{it} \mid X, F, \Lambda] = 0$.

The $r$ interactive terms $\lambda_{ik} F_{kt}$ generalise both additive
unit effects ($r=1$, $F_t = 1$) and additive time effects ($r=1$,
$\lambda_i = 1$).  With $r$ unrestricted, (1) subsumes TWFE as the
special case $r = 0$.

## What `xtife` provides

`xtife` is a pure base-R package (no compiled code, no external
dependencies) offering:

| Feature | Function | Reference |
|---|---|---|
| Balanced IFE: analytical SEs | `ife()` | Bai (2009) |
| Balanced IFE: static bias correction | `ife(..., bias_corr=TRUE)` | Bai (2009, §7) |
| Balanced IFE: dynamic regressors | `ife(..., method="dynamic")` | Moon & Weidner (2017) |
| Factor number selection (balanced) | `ife_select_r()` | Bai & Ng (2002); Bai (2009) |
| Unbalanced IFE: estimation & analytical inference | `ife_unbalanced()` | Su, Wang & Wang (2025); Bai (2009); Bai & Ng (2021) |
| Unbalanced IFE: additive FE (`force`) | `ife_unbalanced(force=)` | Su, Wang & Wang (2025) |
| Unbalanced IFE: factor selection | `ife_select_r_unb()` | Su, Wang & Wang (2025); Bai & Ng (2021) |

---

# Balanced Panel IFE

## Data

The package includes the `cigar` dataset: 46 US states $\times$ 30 years
(1963--1992) of cigarette sales and prices from Baltagi (1995).

```{r data}
library(xtife)
data(cigar)
dim(cigar)
head(cigar[, c("state", "year", "sales", "price")], 3)
```

The panel is balanced: every state appears in every year.

## Quick start

```{r quickstart}
fit <- ife(sales ~ price,
           data  = cigar,
           index = c("state", "year"),
           r     = 2,
           force = "two-way",
           se    = "cluster")
print(fit)
```

The `force` argument controls additive fixed effects:
`"none"` | `"unit"` | `"time"` | `"two-way"` (default).

Key output fields:

```{r components}
fit$coef          # named coefficient vector
fit$se            # standard errors
fit$ci            # 95% confidence intervals (matrix: coef x 2)
fit$converged     # TRUE if outer loop converged
fit$n_iter        # number of outer iterations
dim(fit$F_hat)    # T x r estimated factors
dim(fit$Lambda_hat) # N x r estimated loadings
```

## Estimation algorithm

The Bai (2009) estimator alternates between two steps until convergence:

**Step 1 — Factor extraction.**  Given $\hat\beta$, form
$W_{it} = \ddot y_{it} - \ddot X_{it}'\hat\beta$ (residuals after additive
FE demeaning).  The top-$r$ eigenvectors of $WW'/NT$ (scaled to satisfy
$\hat F' \hat F / T = I_r$) give $\hat F$.  The loadings follow by
regression: $\hat\Lambda = W'\hat F / T$.

**Step 2 — Coefficient update.**  Project the demeaned regressors onto the
orthogonal complement of $\hat F$:

$$\tilde X_{it} = \ddot X_{it} - \hat F_t (\hat F' \hat F)^{-1} \hat F' \ddot X_i$$

Then $\hat\beta = (\tilde X'\tilde X)^{-1}\tilde X'\ddot y$ (Frisch-Waugh-Lovell).

The outer loop converges when $\max|\hat\beta^{(k+1)} - \hat\beta^{(k)}| < \text{tol}$
(default $10^{-9}$).

## Standard error types

```{r se_compare}
fit_std <- ife(sales ~ price, data = cigar,
               index = c("state", "year"), r = 2, se = "standard")
fit_rob <- ife(sales ~ price, data = cigar,
               index = c("state", "year"), r = 2, se = "robust")
fit_cl  <- ife(sales ~ price, data = cigar,
               index = c("state", "year"), r = 2, se = "cluster")

data.frame(
  se_type = c("standard", "robust (HC1)", "cluster"),
  coef    = round(c(fit_std$coef, fit_rob$coef, fit_cl$coef), 5),
  se      = round(c(fit_std$se,   fit_rob$se,   fit_cl$se),   5),
  t_stat  = round(c(fit_std$tstat, fit_rob$tstat, fit_cl$tstat), 3)
)
```

All three use the sandwich formula
$\widehat{\mathrm{Var}}(\hat\beta) = A^{-1} B A^{-1}$
where $A = \tilde X'\tilde X$ and:

| `se=` | $B$ matrix | Degrees of freedom |
|---|---|---|
| `"standard"` | $\hat\sigma^2 A$ | $NT - p - r(N+T-r) - \text{FE dof}$ |
| `"robust"` | $\sum_{it} \hat u_{it}^2 \tilde x_{it}\tilde x_{it}'$ (HC1) | same |
| `"cluster"` | $\sum_i \psi_i \psi_i'$, $\psi_i = \sum_t \hat u_{it}\tilde x_{it}$ | same |

**Rule of thumb**: use `"cluster"` for most macro panels; `"robust"` when
clustering by unit is not appropriate; `"standard"` only as a benchmark.

## Comparison with TWFE

Setting $r = 0$ reduces `ife()` to the standard TWFE estimator:

```{r twfe_compare}
fit0 <- ife(sales ~ price, data = cigar,
            index = c("state", "year"), r = 0, force = "two-way")

# Manual within-transformation
cigar$y_dm <- cigar$sales - ave(cigar$sales, cigar$state) -
                             ave(cigar$sales, cigar$year) + mean(cigar$sales)
cigar$x_dm <- cigar$price - ave(cigar$price, cigar$state) -
                             ave(cigar$price, cigar$year) + mean(cigar$price)
lm0 <- lm(y_dm ~ x_dm - 1, data = cigar)

cat(sprintf("ife (r=0): %.7f\n", fit0$coef["price"]))
cat(sprintf("lm TWFE:  %.7f\n", coef(lm0)["x_dm"]))
cat(sprintf("diff:     %.2e\n",
            abs(fit0$coef["price"] - coef(lm0)["x_dm"])))
```

The difference is at machine precision, confirming algebraic equivalence.
TWFE gives $\hat\beta \approx -0.38$; IFE with $r=2$ gives $\approx -0.52$:
the gap reflects the OVB from ignoring interactive confounders.

## Factor number selection

`ife_select_r()` fits models for $r = 0, \ldots, r_{\max}$ and reports
five information criteria.

```{r select_r, eval=FALSE}
# Not run during build (takes ~20 s); set verbose=FALSE in scripts
sel <- ife_select_r(sales ~ price, data = cigar,
                    index = c("state", "year"),
                    r_max = 6, force = "two-way", verbose = FALSE)
attr(sel, "suggested")   # named vector: r chosen by each IC
```

The five criteria:

| Criterion | Penalty | Notes |
|---|---|---|
| IC1 | $\hat\sigma^2 (N+T-r)\log(NT)/(NT)$ | Bai & Ng (2002) |
| IC2 | $\hat\sigma^2 (N+T-r)\log(\min(N,T)^2)/(NT)$ | Bai & Ng (2002) |
| IC3 | $\hat\sigma^2 (N+T-r)\log(\min(N,T))/(NT)$ | Bai & Ng (2002) |
| IC(BIC) | $V(r)\log(NT)/(NT)$ | Bai (2009, §4.3) |
| PC | $V(r) + r \cdot \hat\sigma^2 (N+T)\log(NT)/(NT)$ | Bai (2009) |

**Recommendation**: IC(BIC) for small-to-moderate panels
($\min(N,T) < 60$); IC1 or IC2 for large panels.

## Asymptotic bias correction (balanced panel)

### Static regressors — Bai (2009)

The IFE estimator has asymptotic bias $O(1/N + 1/T)$ from the incidental
parameters in $\hat\Lambda$ and $\hat F$.  The analytical correction is:

$$\hat\beta^\dagger = \hat\beta - \hat B / N - \hat C / T$$

where $\hat B$ and $\hat C$ are closed-form expressions involving
$\hat F$, $\hat\Lambda$, and the regression residuals (Bai 2009, Theorem 2).

```{r bias_static}
fit_bc <- ife(sales ~ price, data = cigar,
              index = c("state", "year"), r = 2,
              se = "standard", bias_corr = TRUE)
cat(sprintf("Raw IFE:  %.5f\n", fit_bc$coef_raw["price"]))
cat(sprintf("Corrected:%.5f\n", fit_bc$coef["price"]))
cat(sprintf("B_hat:    %.5f  (B_hat/N = %.5f)\n",
            fit_bc$B_hat["price"], fit_bc$B_hat["price"] / fit_bc$N))
cat(sprintf("C_hat:    %.5f  (C_hat/T = %.5f)\n",
            fit_bc$C_hat["price"], fit_bc$C_hat["price"] / fit_bc$T))
```

The correction is most important when $T/N$ is non-negligible.
For the cigar panel ($N=46$, $T=30$, $T/N \approx 0.65$) it shifts the
coefficient by about 0.007.

### Dynamic regressors — Moon and Weidner (2017)

When regressors include lagged outcomes or are otherwise only *weakly*
exogenous (correlated with past errors), the static bias correction is
insufficient.  The `method = "dynamic"` option applies the Moon and
Weidner (2017) double-projection algorithm and three-term bias correction:

$$\hat\beta^* = \hat\beta +
\hat W_X^{-1}\!\left(\frac{\hat B_1}{T} + \frac{\hat B_2}{N}
+ \frac{\hat B_3}{T}\right)$$

$\hat B_1$ (Nickell-type) is near zero for approximately exogenous
regressors.

```{r bias_dynamic}
fit_dyn <- ife(sales ~ price, data = cigar,
               index = c("state", "year"), r = 2,
               se = "standard", method = "dynamic",
               bias_corr = TRUE, M1 = 1L)
cat(sprintf("Dynamic BC coef: %.5f\n", fit_dyn$coef["price"]))
cat(sprintf("B1/T (Nickell):  %.5f\n",
            fit_dyn$B1_hat["price"] / fit_dyn$T))
```

For `price` (approximately exogenous), $\hat B_1 / T \approx 0$, so the
static and dynamic corrected estimates nearly coincide.

---

# Unbalanced Panel IFE

## Motivation

Real-world panels are rarely balanced.  Survey non-response, merger events,
data-collection gaps, and rotating samples all cause missing $(i,t)$ cells.
Simply deleting units with incomplete records (list-wise deletion) induces
selection bias; imputation by column means distorts the factor structure.

`ife_unbalanced()` extends the Bai (2009) interactive fixed effects
estimator to genuinely unbalanced panels, following the estimation and
inference theory of Su, Wang and Wang (2025), and combines three published
ideas:

1. **Factor analysis with missing data.** The factor structure is recovered
   from the observed cells by an expectation-maximisation (EM) algorithm that
   treats the unobserved entries as latent, in the spirit of Bai (2009,
   Appendix B) and the missing-data factor / matrix-completion framework of
   Bai and Ng (2021).
2. **Nuclear-norm initialisation.** A nuclear-norm-regularised (soft-impute)
   estimator (Mazumder, Hastie and Tibshirani 2010) provides a consistent
   warm start for the factor structure, stabilising the alternating loop when
   the panel is severely unbalanced.
3. **Analytical inference.** Standard errors use a sandwich estimator on
   factor-projected regressors (the unbalanced analogue of the balanced
   formula of Bai 2009), with heteroskedasticity-, cluster- and
   autocorrelation-robust variants; an analytical incidental-parameter bias
   correction is provided for both strictly exogenous and predetermined
   (weakly exogenous) regressors. The inference and bias-correction formulae
   for the unbalanced case follow Su, Wang and Wang (2025), extending the
   bias-correction principles of Bai (2009) and Moon and Weidner (2017).

In addition, additive unit and/or time fixed effects can be removed jointly
with the factors through the `force` argument (Section 3.2).

## The unbalanced IFE model

Let $d_{it} \in \{0,1\}$ indicate whether $(i,t)$ is observed.  The model is:

$$y_{it} = X_{it}'\beta + \lambda_i' F_t + u_{it}
\qquad \text{for all } (i,t) \text{ with } d_{it}=1 \qquad (2)$$

By default (`force = "none"`) there are no additive fixed effects in (2):
individual and time heterogeneity is absorbed entirely by the interactive
factor structure. Additive unit and/or time effects can instead be modelled
explicitly through the `force` argument (see below), which is the recommended
route when the data contain a strong common level or trend.

### Identification

The normalisation $\hat F' \hat F / T = I_r$ and
$\hat\Lambda' \hat\Lambda$ diagonal identify $F$ and $\Lambda$ up to an
$r \times r$ rotation.  Only the column space of $F$ (the factor space)
is identified, which suffices for inference on $\beta$.

### Additive fixed effects via `force`

The `force` argument adds additive unit ($\alpha_i$) and/or time ($\xi_t$)
effects to model (2):

$$y_{it} = \alpha_i + \xi_t + X_{it}'\beta + \lambda_i' F_t + u_{it}.$$

Setting `force = "unit"`, `"time"`, or `"two-way"` estimates these effects
*jointly* with the factors and slope, by removing them from the imputed
(completed) panel inside the EM loop. Because the demeaning is performed on
the imputed --- rather than the observed-only --- panel, it is robust to
informative (selection-related) missingness, unlike a naïve
observed-cell demeaning. The default `force = "none"` (intercept-free) is
appropriate when level/trend structure is already captured by the factors;
choose `force = "two-way"` (matching the balanced `ife()` default) for data
with a pronounced common level or trend. The degrees-of-freedom adjustment
for the additive parameters ($N-1$, $T-1$, or $N+T-2$) is carried through to
the standard errors automatically.

## Quick start with unbalanced data

```{r unbalanced_setup}
# Create a 10% randomly missing version of the cigar panel
set.seed(42)
cigar_unb <- cigar[sample(nrow(cigar), size = floor(0.9 * nrow(cigar))), ]
cat(sprintf("Observations: %d / %d  (%.0f%% fill)\n",
            nrow(cigar_unb), nrow(cigar), 100 * nrow(cigar_unb) / nrow(cigar)))
```

```{r unbalanced_fit}
fit_unb <- ife_unbalanced(sales ~ price,
                           data  = cigar_unb,
                           index = c("state", "year"),
                           r     = 2L,
                           se    = "cluster")
print(fit_unb)
```

Key output fields mirror those of `ife()`:

```{r unbalanced_fields}
fit_unb$coef       # estimated beta
fit_unb$se         # standard errors
fit_unb$ci         # 95% confidence intervals
fit_unb$converged  # outer-loop convergence
fit_unb$n_obs      # number of observed (i,t) pairs
fit_unb$N          # number of units
fit_unb$TT         # number of distinct time periods
dim(fit_unb$F_hat)       # TT x r
dim(fit_unb$Lambda_hat)  # N  x r
```

## Estimation algorithm

### Outer alternating-maximisation (AM) loop

`ife_unbalanced()` solves the unbalanced least-squares problem:

$$(\hat\beta, \hat F, \hat\Lambda) = \arg\min_{\beta, F, \Lambda}
\sum_{i,t} d_{it}\left(y_{it} - X_{it}'\beta - \lambda_i' F_t\right)^2$$

subject to $F'F/T = I_r$.

The outer loop alternates:

1. **Factor step**: given $\hat\beta$, form $W_{it} = y_{it} - X_{it}'\hat\beta$
   and apply the EM inner loop (below) to recover $(\hat F, \hat\Lambda)$.
2. **Coefficient step**: project out the factor space from each unit's
   observed data and run pooled OLS on the projected regressors
   $\tilde X_{it} = X_{it} - \hat F_t(\hat F_i'\hat F_i)^{-1}\hat F_i' X_i$,
   where the inner product is restricted to unit $i$'s observed periods.

### EM inner loop (Bai 2009, Appendix B) — used within each AM iteration

Within each AM outer iteration, $(\hat F, \hat\Lambda)$ are updated by the
EM algorithm of Bai (2009, Appendix B), which treats the missing $W_{it}$
as latent variables imputed from the current factor estimate:

1. **E-step**: impute missing cells with the current fitted values:
   $W_{it}^{\text{fill}} = \hat\Lambda_i'\hat F_t$ for all unobserved $(i,t)$.
2. **M-step**: apply standard SVD factor extraction to the completed
   $T \times N$ matrix $W^{\text{fill}}$.

The EM loop converges when $\max_{(i,t): d_{it}=1}
|\hat\Lambda_i^{(k+1)}\!\!'\hat F_t^{(k+1)} - \hat\Lambda_i^{(k)}\!\!'\hat F_t^{(k)}|
< \text{tol}_{\text{EM}}$ (convergence on fitted values at observed cells;
the rotational indeterminacy of $F$ makes direct comparison of $\hat F$
meaningless).

### Initialisation

Two options for the initial $(\hat\beta^{(0)}, \hat F^{(0)}, \hat\Lambda^{(0)})$:

**`init = "ols"` (default):** plain OLS on observed cells gives
$\hat\beta^{(0)}$; the EM inner loop is then run from $(\hat F, \hat\Lambda) = 0$.

**`init = "nnr"` (nuclear-norm regularisation):** solves the nuclear-norm
penalised problem

$$(\hat\beta^{(0)}, \hat\Theta^{(0)}) = \arg\min_{\beta, \Theta}
\tfrac{1}{2}\!\sum_{it} d_{it}(y_{it} - \Theta_{it} - X_{it}'\beta)^2
+ \nu_{NT}\|\Theta\|_*$$

via soft-impute iteration.  The penalty $\nu_{NT}$ is cross-validated over
$c \cdot \sqrt{\max(N,T)}$ with $c \in \{0.01, 0.1, 1, 10\}$.

NNR initialisation is recommended when the panel is severely unbalanced
(fill rate $< 60\%$) or has a rich factor structure ($r \geq 3$).  For
mildly unbalanced panels, both initialisations converge to the same fixed
point.

```{r init_compare}
# Both converge to the same coefficient for mild unbalancedness
fit_ols <- ife_unbalanced(sales ~ price, data = cigar_unb,
                           index = c("state", "year"), r = 2, init = "ols")
fit_nnr <- ife_unbalanced(sales ~ price, data = cigar_unb,
                           index = c("state", "year"), r = 2, init = "nnr")
cat(sprintf("OLS init: coef = %.5f,  %d outer iterations\n",
            fit_ols$coef["price"], fit_ols$n_iter))
cat(sprintf("NNR init: coef = %.5f,  %d outer iterations\n",
            fit_nnr$coef["price"], fit_nnr$n_iter))
```

## Analytical standard errors

The sandwich variance of $\hat\beta$ follows Su, Wang and Wang (2025) and
requires projecting each regressor out of both the factor space and the
loading space, giving *doubly-projected* regressors $\hat x_{itk}$ (the
unbalanced analogue of the Frisch--Waugh projection used in the balanced
case):

$$\hat x_{itk} = x_{itk} - \hat\delta_{ki}'\hat F_t - \hat\omega_{kt}'\hat\lambda_i$$

where $\hat\delta_{ki}$ and $\hat\omega_{kt}$ solve, over the observed cells,
the alternating least-squares system

$$\hat\delta_{ki} = \Bigl(\sum_t d_{it}\hat F_t\hat F_t'\Bigr)^{-1}
\!\sum_t d_{it}\hat F_t(x_{itk} - \hat\lambda_i'\hat\omega_{kt})$$

$$\hat\omega_{kt} = \Bigl(\sum_i d_{it}\hat\lambda_i\hat\lambda_i'\Bigr)^{-1}
\!\sum_i d_{it}\hat\lambda_i(x_{itk} - \hat F_t'\hat\delta_{ki})$$

The sandwich variance is then

$$\widehat{\mathrm{Var}}(\hat\beta) = \hat W_X^{-1} \hat\Omega_X \hat W_X^{-1}$$

with $\hat W_X = \frac{1}{NT}\sum_{it} d_{it}\hat x_{it}\hat x_{it}'$ and
$\hat\Omega_X$ depending on the `se` argument:

| `se=` | $\hat\Omega_X$ | Assumption |
|---|---|---|
| `"standard"` | $\frac{1}{NT}\sum_{it} d_{it}\hat u_{it}^2\hat x_{it}\hat x_{it}'$ | Homoskedastic errors |
| `"robust"` | same formula, HC1-scaled | Heteroskedastic errors |
| `"cluster"` | $\frac{1}{NT}\sum_i \psi_i\psi_i'$, $\psi_i = \sum_t d_{it}\hat u_{it}\hat x_{it}$ | Serial correlation within units |
| `"hac"` | Bartlett kernel over $(t,s)$ pairs: $\frac{1}{NT}\sum_i\sum_{t,s}\Gamma\!\left(\frac{|s-t|}{L_T}\right)d_{it}d_{is}\hat u_{it}\hat u_{is}\hat x_{it}\hat x_{is}'$ | Serial correlation over time within units |

The cluster-robust variant follows the within-unit clustering of Arellano
(1987) and Cameron, Gelbach and Miller (2011); the `"hac"` variant uses the
Bartlett kernel of Newey and West (1987),
$\Gamma(u) = (1-u)\mathbf{1}\{u \leq 1\}$, with bandwidth defaulting to
$L_T = \lfloor 2T^{1/5}\rfloor$.

```{r se_unb}
fit_std <- ife_unbalanced(sales ~ price, data = cigar_unb,
                           index = c("state", "year"), r = 2, se = "standard")
fit_rob <- ife_unbalanced(sales ~ price, data = cigar_unb,
                           index = c("state", "year"), r = 2, se = "robust")
fit_cl  <- ife_unbalanced(sales ~ price, data = cigar_unb,
                           index = c("state", "year"), r = 2, se = "cluster")
fit_hac <- ife_unbalanced(sales ~ price, data = cigar_unb,
                           index = c("state", "year"), r = 2, se = "hac")

data.frame(
  se_type = c("standard", "robust", "cluster", "hac"),
  coef    = round(c(fit_std$coef, fit_rob$coef, fit_cl$coef, fit_hac$coef), 5),
  se      = round(c(fit_std$se,   fit_rob$se,   fit_cl$se,   fit_hac$se),   5)
)
```

**When to use `"hac"`**: when the errors $u_{it}$ are serially correlated
(e.g., in dynamic models or when residuals show autocorrelation).  For
strictly exogenous regressors with i.i.d. errors, `"cluster"` is sufficient.

## Factor number selection

For unbalanced panels, `ife_select_r_unb()` uses the singular value
thresholding (SVT) rule of Su, Wang and Wang (2025) applied to the
nuclear-norm regularised matrix $\hat\Theta^{(0)}$ --- a missing-data
counterpart of the eigenvalue-ratio / information-criterion rules of Bai and
Ng (2002):

$$\hat r = \sum_{s=1}^{N \wedge T}
\mathbf{1}\!\left\{
\sigma_s\!\left(\frac{\hat\Theta^{(0)}}{\sqrt{NT}}\right) \geq
c_f\sqrt{c_{NT}^{-1/4}\,\sigma_1\!\left(\frac{\hat\Theta^{(0)}}{\sqrt{NT}}\right)}
\right\}$$

where $c_{NT} = \min(\sqrt{N}, \sqrt{T})$ and $c_f = 0.6$ is the default
threshold constant, which performs well for moderate degrees of
unbalancedness.

```{r sel_unb, eval=FALSE}
# Not run during build (NNR cross-validation takes ~10 s)
sel_unb <- ife_select_r_unb(sales ~ price, data = cigar_unb,
                              index = c("state", "year"), verbose = FALSE)
cat(sprintf("SVT selects r_hat = %d\n", sel_unb$r_hat))
```

The function returns (invisibly):

- `r_hat`: the selected number of factors (at least 1);
- `sv`: the normalised singular values used in the threshold comparison;
- `threshold`: the computed threshold value;
- `nu_used`: the NNR penalty chosen by cross-validation.

## Analytical bias correction

Even with the correct $r$, the unbalanced IFE estimator carries an
incidental-parameter bias of order $O(N^{-1}) + O(T^{-1})$, analogous to the
$B/N + C/T$ bias of the balanced estimator (Bai 2009) and the dynamic bias of
Moon and Weidner (2017). Setting `bias_corr = TRUE` applies the analytical
correction of Su, Wang and Wang (2025),
$\hat\beta^{abc} = \hat\beta - (NT)^{-1/2}\hat W_X^{-1}\hat b$,
where, for strictly exogenous regressors,
$\hat b = \hat b_3 + \hat b_4 + \hat b_5 + \hat b_6$ with

$$\hat b_{3k} = \frac{1}{\sqrt{NT}}\sum_{i,t}
d_{it}\hat u_{it}^2 \hat\delta_{ki}'[\hat L_{ff}^{-1}]_t\hat\lambda_i$$

$$\hat b_{4k} = \frac{1}{\sqrt{NT}}\sum_{i,t}
d_{it}\hat u_{it}^2 \hat\omega_{kt}'[\hat L_{\lambda\lambda}^{-1}]_i\hat F_t$$

$$\hat b_{5k} = \frac{1}{\sqrt{NT}}\sum_{i,t}
d_{it}\hat u_{it}^2 \hat\lambda_i'\hat\Xi_{kt}\hat\lambda_i$$

$$\hat b_{6k} = \frac{1}{\sqrt{NT}}\sum_{i,t}
d_{it}\hat u_{it}^2 \hat F_t'\hat\Delta_{ki}\hat F_t$$

Here $[\hat L_{ff}^{-1}]_t = (\sum_i d_{it}\hat\lambda_i\hat\lambda_i')^{-1}$,
$[\hat L_{\lambda\lambda}^{-1}]_i = (\sum_t d_{it}\hat F_t\hat F_t')^{-1}$,
and $\hat\Delta_{ki}$, $\hat\Xi_{kt}$ are $r \times r$ auxiliary matrices
formed from $\hat\delta$, $\hat\omega$ and the above inverses; $\hat u_{it}$
are the model residuals. Full expressions are given in the
`.ife_bias_unb()` source documentation.

### Strictly exogenous regressors (`exog = "strict"`)

Use when $\mathbb{E}[u_{it} \mid X, F, \Lambda] = 0$ for all $(i,t)$.
This is the default and covers the vast majority of cross-sectional
panel applications.

```{r bc_strict}
fit_bc_strict <- ife_unbalanced(sales ~ price, data = cigar_unb,
                                 index = c("state", "year"), r = 2,
                                 se = "standard", bias_corr = TRUE,
                                 exog = "strict")
cat(sprintf("Raw beta:       %.5f\n", fit_bc_strict$coef_raw["price"]))
cat(sprintf("Corrected beta: %.5f\n", fit_bc_strict$coef["price"]))
cat(sprintf("b_hat (b3+...+b6): %.5f\n", fit_bc_strict$b_hat["price"]))
```

### Weakly exogenous regressors (`exog = "weak"`)

Use when the regressor includes the lagged dependent variable or is
otherwise predetermined (correlated with past but not future errors).
This adds the dynamic bias term $\hat b_2$:

$$\hat b_{2k} = \frac{1}{\sqrt{NT}}\sum_i\sum_{t < s}
\Gamma\!\left(\frac{s-t}{L_T}\right)
\hat u_{it}\, x_{isk}\,\hat F_t'[\hat L_{\lambda\lambda}^{-1}]_i\hat F_s$$

The total bias correction becomes
$\hat b = \hat b_2 + \hat b_3 + \hat b_4 + \hat b_5 + \hat b_6$.

```{r bc_weak, eval=FALSE}
# Not run: exog="weak" adds the b2 dynamic term
# (requires HAC SE for valid inference in dynamic models)
fit_bc_weak <- ife_unbalanced(y ~ lag_y + x, data = df_dynamic,
                               index = c("unit", "time"), r = 2,
                               se = "hac", bias_corr = TRUE,
                               exog = "weak")
```

**Summary of `exog` × `se` combinations:**

| Regressor type | Errors | Recommended | Bias term |
|---|---|---|---|
| Strictly exogenous | i.i.d. | `se="standard"` or `"cluster"` | $b_3$–$b_6$ |
| Strictly exogenous | Serially correlated | `se="hac"` | $b_3$–$b_6$ (HAC versions) |
| Weakly exogenous (lag $y$) | i.i.d. | `se="cluster"` + `exog="weak"` | $b_2$–$b_6$ |
| Weakly exogenous (lag $y$) | Serially correlated | `se="hac"` + `exog="weak"` | $b_2$–$b_6$ (HAC) |

## Balanced vs unbalanced on the same data

As a sanity check, `ife_unbalanced()` on a complete (balanced) panel
should agree with `ife()` under the same `force` specification. We use a
simulated mean-zero factor panel (on data with a large level, such as
`cigar`, `force = "none"` is ill-conditioned for both estimators -- the
factors must absorb the level -- so use `force = "two-way"` there):

```{r bal_vs_unbal}
set.seed(99)
N <- 30; TT <- 22
F0 <- matrix(rnorm(TT * 2), TT, 2); L0 <- matrix(rnorm(N * 2), N, 2)
X  <- matrix(rnorm(N * TT), TT, N)
Y  <- F0 %*% t(L0) + 0.7 * X + matrix(rnorm(N * TT), TT, N)
sim <- data.frame(unit = rep(seq_len(N), each = TT),
                  time = rep(seq_len(TT), times = N),
                  Y = as.vector(Y), X = as.vector(X))

fit_bal <- ife(Y ~ X, data = sim, index = c("unit", "time"),
               r = 2, force = "none", se = "standard")

fit_unb2 <- ife_unbalanced(Y ~ X, data = sim, index = c("unit", "time"),
                            r = 2, force = "none", se = "standard")

cat(sprintf("ife (balanced):     %.6f\n", fit_bal$coef["X"]))
cat(sprintf("ife_unbalanced:     %.6f\n", fit_unb2$coef["X"]))
cat(sprintf("Difference:         %.2e\n",
            abs(fit_bal$coef["X"] - fit_unb2$coef["X"])))
```

The two estimators agree closely (they follow different iteration paths,
so agreement is to roughly $10^{-2}$ rather than machine precision).

---

# Input validation and common errors

Both `ife()` and `ife_unbalanced()` perform strict input validation and
stop with an informative message if any check fails.

| Check | Error message |
|---|---|
| `data` is not a `data.frame` | `"'data' must be a data.frame"` |
| index columns not in `data` | `"'index' columns not found: ..."` |
| NA in outcome or covariates | `"Missing values (NA) in variable '...'"` |
| Duplicate (unit, time) pairs | `"Duplicate (unit, time) pairs found"` |
| `r` < 1 (unbalanced) | `"'r' must be a positive integer"` |
| Panel unbalanced in `ife()` | `"Panel is not balanced: ..."` |
| `se` not in allowed set | `"'se' must be one of: ..."` |
| `exog` not `"strict"`/`"weak"` | `"'exog' must be 'strict' or 'weak'"` |
| `L_T < 1` | `"'L_T' must be a positive integer"` |

---

# Quick reference: argument tables

## `ife()` — balanced panel

| Argument | Default | Description |
|---|---|---|
| `formula` | — | `outcome ~ covariate1 + ...` |
| `data` | — | balanced long-format `data.frame` |
| `index` | — | `c("unit_col", "time_col")` |
| `r` | `1L` | number of factors |
| `force` | `"two-way"` | additive FE: `"none"` / `"unit"` / `"time"` / `"two-way"` |
| `se` | `"standard"` | `"standard"` / `"robust"` / `"cluster"` |
| `method` | `"static"` | `"static"` / `"dynamic"` (Moon & Weidner 2017) |
| `bias_corr` | `FALSE` | apply analytical bias correction |
| `M1` | `1L` | MA order for dynamic BC (used when `method="dynamic"`) |
| `tol` | `1e-9` | outer-loop convergence tolerance |
| `max_iter` | `10000L` | outer-loop iteration cap |

## `ife_unbalanced()` — unbalanced panel

| Argument | Default | Description |
|---|---|---|
| `formula` | — | `outcome ~ covariate1 + ...` |
| `data` | — | long-format `data.frame` (may be unbalanced) |
| `index` | — | `c("unit_col", "time_col")` |
| `r` | `1L` | number of factors |
| `force` | `"none"` | additive FE: `"none"` / `"unit"` / `"time"` / `"two-way"` (estimated jointly with the factors; default is intercept-free) |
| `se` | `"standard"` | `"standard"` / `"robust"` / `"cluster"` / `"hac"` |
| `init` | `"ols"` | initialisation: `"ols"` or `"nnr"` (nuclear-norm) |
| `bias_corr` | `FALSE` | apply analytical incidental-parameter bias correction |
| `exog` | `"strict"` | exogeneity: `"strict"` or `"weak"` |
| `L_T` | `NULL` | HAC bandwidth ($\lfloor 2T^{1/5}\rfloor$ if `NULL`) |
| `c_f` | `0.6` | SVT threshold constant (used with `ife_select_r_unb()`) |
| `nu_NT` | `NULL` | NNR penalty (cross-validated if `NULL`) |
| `tol` | `1e-9` | outer-loop convergence tolerance |
| `max_iter` | `10000L` | outer-loop iteration cap |
| `tol_em` | `1e-7` | EM inner-loop convergence tolerance |
| `max_iter_em` | `500L` | EM inner-loop iteration cap |

---

# References

Bai, J. (2009). Panel data models with interactive fixed effects.
*Econometrica*, **77**(4), 1229--1279.
<doi:10.3982/ECTA6135>

Bai, J. and Ng, S. (2002). Determining the number of factors in approximate
factor models.
*Econometrica*, **70**(1), 191--221.
<doi:10.1111/1468-0262.00273>

Bai, J. and Ng, S. (2021). Matrix completion, counterfactuals, and factor
analysis of missing data.
*Journal of the American Statistical Association*, **116**(536), 1746--1763.
<doi:10.1080/01621459.2021.1967163>

Baltagi, B. H. (1995). *Econometric Analysis of Panel Data*. Wiley.

Arellano, M. (1987). Computing robust standard errors for within-groups
estimators.
*Oxford Bulletin of Economics and Statistics*, **49**(4), 431--434.

Cameron, A. C., Gelbach, J. B. and Miller, D. L. (2011). Robust inference with
multiway clustering.
*Journal of Business & Economic Statistics*, **29**(2), 238--249.
<doi:10.1198/jbes.2010.07136>

Mazumder, R., Hastie, T. and Tibshirani, R. (2010). Spectral regularization
algorithms for learning large incomplete matrices.
*Journal of Machine Learning Research*, **11**, 2287--2322.

Moon, H. R. and Weidner, M. (2017). Dynamic linear panel regression models
with interactive fixed effects.
*Econometric Theory*, **33**, 158--195.
<doi:10.1017/S0266466615000328>

Newey, W. K. and West, K. D. (1987). A simple, positive semi-definite,
heteroskedasticity and autocorrelation consistent covariance matrix.
*Econometrica*, **55**(3), 703--708.
<doi:10.2307/1913610>

Su, L., Wang, F. and Wang, Y. (2025). Estimation and inference for interactive
fixed effects panel data models with unbalanced panels.
SSRN Working Paper No. 5177283.
<doi:10.2139/ssrn.5177283>
